---
title: "Lab09"
author: "Catherina Santoso (PID: A18552334)"
format: pdf
---

In this mini-project, you will explore 538's Halloween Candy dataset.

We will use los of **ggplot** some basic stats, correlation analysis and PCA to make sense of the landscape of US acndy - something hopefully more relaable than 

## Data Import
Our dataset is a CSV file so we use `read.csv()` and have a wee peak with `head()`

```{r}
candy <- read.csv("candy-data.csv", row.names = 1)
head(candy)

```

>Q1. How many different candy types are in the dataset?

There are `r nrow(candy)` rows in this dataset

```{r}
length(names(candy))

nrow(candy)
```

>Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

>Q3. What is your favorite candy (other than Twix) in the dataset in the dataset and what is it's winpercent value?

```{r}
library(dplyr)

candy |>
  filter(row.names(candy) == "Kit Kat") |>
  select(winpercent)
```

>Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy |>
  filter(row.names(candy) == "Kit Kat") |>
  select(winpercent)
```


>Q5. What is the winpercent value for "Tootsie Roll Snack Bars"?

```{r}
candy |>
  filter(row.names(candy) == "Tootsie Roll Snack Bars") |>
  select(winpercent)
```

>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

```{r}
library("skimr")
skim(candy)
```


>Q7. What do you think a zero and one represent for the `candy$chocolate` column?

```{r}
candy$chocolate

# 1 for the candy contains chocolate, and 0 for candy does not contain chocolate
```

## Exploratory analysis

>Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)

hist(candy$winpercent)

ggplot(candy, aes(x = winpercent)) +
  geom_histogram()
```

>Q9. Is the distribution of winpercent values symmetrical?

No, it's not symmetrical

>Q10. Is the center of the distribution above or below 50%?

```{r}
mean(candy$winpercent)
summary(candy$winpercent)
```
The center is below 50%


>Q11. On average is chocolate candy higher or lower ranked than fruit candy?

1. Find all chocolate candy
2. Get their winpercent values
3. Find the mean
4. Find all fruit candy
5. Get their winpercent values
6. Find the mean
7. Compare the two means

```{r}
chocolate_rank <- candy$winpercent[as.logical(candy$chocolate)]
candy_rank <- candy$winpercent[as.logical(candy$fruity)]

t.test(x = chocolate_rank, y = candy_rank)

```

or

```{r}
choc.candy <- candy[candy$chocolate == 1, ]
choc.win <- choc.candy$winpercent
mean(choc.win)
```

```{r}
fruit.win <- candy[candy$fruity == 1, ]$winpercent
mean(fruit.win)
```
```{r}
t.test(choc.win, fruit.win)
```


>Q12. Is this difference statistically significant?

The difference is statistically significant because the p-value of 2.871e-08 < 0.05.


>Q13. What are the five least liked candy types in this set?

```{r}
y <- c("y", "a", "z")
sort(y)
```

```{r}
y
order(y)
```


```{r}
ord.ind <- order(candy$winpercent)
head(candy[ord.ind,],5)
```


>Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
tail(candy[ord.ind,], 5)
```

>Q15. Make a first barplot of candy ranking based on winpercent values.

We need a custom color vector

```{r}
my_cols <- rep("black", nrow(candy))
my_cols[candy$chocolate == 1] <- "chocolate"
my_cols[candy$fruity == 1] <- "brown"
my_cols[candy$bar == 1] <- "pink"
my_cols
```

```{r}
ggplot(candy) + 
  aes(winpercent,
      reorder(row.names(candy), winpercent) ) +
  geom_col() +
  ylab("")
```

```{r}
ggplot(candy) + 
  aes(winpercent,
      reorder(row.names(candy), winpercent)) +
  geom_col(col = my_cols) +
  ylab("")
```


## Taking a look at pricepercent
```{r}
library(ggrepel)

# How about a plot of win vs price
ggplot(candy) +
  aes(x = winpercent, y = pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) +
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)

```

>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

```{r}
exp_ord <- order(candy$pricepercent)
head( candy[exp_ord,c(11,12)], n=5 )
```
The candy that offers the most bang for your buck is Tootsie Roll Midgies, which is cheaper than 99% of candies, with a winpercent of 45.7%.

>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```
The most expensive candy in the dataset is Nik L Nip, which is more expensive than 97.6% of candies, and having a low winpercent of 22.44%.

## Exploring the correlation structure
```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```

>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate/fruity and pariblus/bar variables are anti-correlated.

>Q23. Similarly, what two variables are most positively correlated?

Chocolate and bar are the mos positively correlated, because most bars have chocolate.


## Principal Component Analysis

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(size = 3, col = my_cols) + 
        geom_text(size = 3, col = my_cols) 

p
```

```{r}
ggplot(pca$rotation) +
  aes(PC1, 
  reorder(row.names(pca$rotation), PC1)) +
  geom_col()
```


>Q24. Complete the code to generate the loadings plot above. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? Where did you see this relationship highlighted previously?

Fruity, pluribus, and hard are variables strongly picked up by PC1 in the positive direction. This makes sense to me because hard candies are stored together in jars. We also saw them correlated with one another in previous plots. 


>Q25. Based on your exploratory analysis, correlation findings, and PCA results, what combination of characteristics appears to make a “winning” candy? How do these different analyses (visualization, correlation, PCA) support or complement each other in reaching this conclusion?

Based on my analysis, characteristics that make a "winning" candy are mostly chocolate-based or bars because they have higher win percent values.







