---
title: "Class 5: Data Viz with ggplot"
author: "Catherina Santoso (PID: A18552334)"
format: pdf
---

## Background

There are lot's of ways to make figures in R. These include so-called "base R" graphics (e.g. `plot()`) and tones of add-on packages like **ggplot2**.

For example here we make the same plot with both:


```{r}
plot(cars)
```

First I need to install the package with the command `install.packages()`
>**N.B.** We never run an install cmd in a quarto code chunk or we will end up re-installing packages many many times - which is not what we want!

```{r}
library(ggplot2)
```



ggplot needs:
- data
- aesthetics
- geom(etry)

```{r}
ggplot(cars) +
  aes(x=speed,y=dist) +
  geom_point() 
```
add a line to better show relationship bw speed and dist

```{r}
p <- ggplot(cars) +
  aes(x=speed,y=dist) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) +
  labs(title="Stopping distance of old cars",
       subtitle = "Data from the `Cars` object",
       x="Speed (mph)",
       y="Distance (ft)")

p
```


```{r}
p + theme_bw()
```

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```


```{r}
ggplot(genes) +
  aes(Condition1, Condition2) +
  geom_point()
```

```{r}
head(genes$State,3)
```

Version 2 let's color by `State` so we can see the up and down significant genes compared to all the unchanging genes

```{r}
ggplot(genes) +
  aes(Condition1, Condition2, col=State) +
  geom_point()
```

Vers 3 plot, modify default color to sth like this
```{r}
ggplot(genes) +
  aes(Condition1, Condition2, col=State) +
  geom_point() +
  scale_color_manual(values=c("purple","pink","blue"))+
  labs(x="Control(no drug)",
       y="Drug",
       title = "Gene Expression Changes upon GLP-1 drug") +
  theme_bw()
```

## Going further

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```


```{r}
head(gapminder,3)
```


```{r}
ggplot(gapminder)+
  aes(x=gdpPercap, y=lifeExp,col=continent)+
  geom_point(alpha=0.3)
```

Let's "facet" (i.e. make a separate plot) by continent rather than the big hot mess above.

```{r}
ggplot(gapminder)+
  aes(x=gdpPercap, y=lifeExp,col=continent)+
  geom_point(alpha=0.3) +
  facet_wrap(~continent)
```





## Custom plots 
How big is this gapminder dataset?
```{r}
nrow(gapminder)
```

I want to "filter" down to a subset of this data. I will use the **dplyr** package to help me.
First I need to install it and then load it up...
`install.packages("dplyr")` and then `library(dplyr)`

```{r}
library(dplyr)

gapminder_2007 <- filter(gapminder,year==2007)
head(gapminder,3)
```

```{r}
filter(gapminder_2007, year==2007, country == "Ireland")

```

> Q. Make a plot comparing 1977 and 2007 for all countries

```{r}
input <- filter(gapminder, year %in% c(1977,2007))
head(input)
```

```{r}

gapminder_1977 <- gapminder %>% filter(year==1977 | year==2007)

ggplot(gapminder_1977) + 
  geom_point(aes(x = gdpPercap, y = lifeExp, color=continent), alpha=0.7) + 
  scale_size_area(max_size = 10) +
  facet_wrap(~year)
```

