---
title: "Class 7: Machine Learning 1"
author: "Catherina Santoso (PID: A18552334)"
format: pdf
---

##  Background
>Today we will begin our exploration of important machine learning methods with a focus on **clustering** and **dimensionality reduction**.

>To start testing these methods let's make up some sample data to cluster where we know what the answer should be.

```{r}
hist( rnorm(3000, mean = 10, sd = ))
```

>Q. Can you generate 30 numbers centered at +3 and 30 numbers centered at -3 taken from a normal distribution?

```{r}
tmp <- c(rnorm(30, mean = 3),
        rnorm(30, mean = -3) )

x <- cbind(x=tmp, y = rev(tmp))
plot(x)
```

## K-means clustering

The main function in "base R" for K-means clustering is called `kmeans()`, let's try it out:

```{r}
k <- kmeans(x, centers = 2)
k
```

>Q. What component of your kmeans result object has the cluster centers?

```{r}
k$centers
```

>Q. What component of your kmeans result object has the cluster size (i.e. how many points are in each cluster)?

```{r}
k$size
```


>Q. What component of your kmeans result object has the cluster membership vector (i.e. the main clustering result: which points are in which cluster)?

```{r}
k$cluster
```


>Q. Plot the results of clustering (i.e. our data colored by the clustering result) along with the cluster centers.

```{r}
plot(x, col = k$cluster)
#points(k$centers, col = "blue", pch = 15, cex =2)
```


>Q. Can you run kmeans again and cluster into 4 clusters and plot the results just like we did above with coloring by cluster and the cluster centers shown in blue?

```{r}
k4 <-  kmeans(x, centers = 4)
plot(x, col = k4$cluster)
points(k$centers, col = "blue", pch = 15, cex =2)

```

> **Key-point**: Kmeans will always return the clustering that we ask for (this is the "K" or "centers" in K-means)!


```{r}
k$tot.withinss
```
## Hierarchial clustering 

The main function for Hierarchical clustering in base R is called `hclust()`. One of the main differnces with respect to the `kmeans()` function is that you can not just pass your input data directly to `hclust()`- it needs a "distance matrix" as input. We can get this from lot's of places including the `dist()` function. 

```{r}
d<-dist(x)
hc<-hclust(d)
plot(hc) 
```

We can "cut" the dendrogram or "tree" at a given height to yield our "clusters". For this we use the function `cutree`
```{r}
plot(hc)
abline(h=10, col="red")
grps <- cutree(hc, h=10)
```

```{r}
grps
```

> Q. Plot our data `x` colored by the clusterinv result from `hclust()` and `cutree`?

```{r}
grps <- cutree (hc, h=10)
plot(x, col=grps)
```

## Principal component Analysis (PCA) 

PCA is a popular dimensionality reduction technique that is widely used in bioinformatics.

##PCA of UK food 

Read data on food consumption in the UK

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
x
```

It looks like the row names are not set properly. We can fix this

```{r}
rownames (x) <- x[,1]
x <-x[,-1]
x
```

```{r}
x <- read.csv(url, row.names = 1)
x
```

>Q. How many rows and cols are in your new daa frame x? What r functions could you use to answer this Q?

```{r}
dim(x)
```


>Q. Which approach to solving the 'row-names problem' mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?

```{r}
row.names(x) # This is better because it fixes on the designated colmun while the x <- x[,-1] function takes away a column from the data
```

>Q. Changing what optional argument in the above barplot() function results in the following plot?

```{r}
barplot(as.matrix(x), beside = T, col = rainbow(nrow(x)))
```


```{r}
barplot(as.matrix(x), beside = F, col = rainbow(nrow(x)))
```

## Pairs plot and heatmaps

>Q4. is missing

>Q5. We can use the pairs() function to generate all pairwise plots for our countries. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagnoal for a given plot?

```{r}
pairs(x, col = rainbow(17), pch = 16, cex = 2)
```

### Heatmap

We can install the **pheatmap** package with the `install.packages()` command that we used previously. Remember that we always run this in the console and not a code chunk in our quarto document.
```{r}
library(pheatmap)
pheatmap(x)

```

of all these plot really only the `pairs()` plot was useful. This however took a bit of work to interpret and well of scale and i am looking at much bigger datasets

## PCA the rescue

The main function in `base R` for PCA is called `prcomp()`


```{r}
pca <- prcomp( t(x) )
summary(pca)
```

> Q. How much variance is captured in the first PC?

67.4%

> Q. How many PCs do I need to capture at least 90% of the total variance in the dataset?

Two PCs capture 96.5% of the total variance 

> Q Plot our main PCA result. Folks can call this different things depending on their field of study e.g. "PC plot", "ordienation plot", "score plot, "PC1 vs PC2"

```{r}
attributes(pca)
```

To generate our PCA score plot we want the `pca$x` component of the result object
```{r}
pca$x
```

```{r}
my_cols <- c("orange", "blue", "green", "purple")
plot(pca$x[,1], pca$x[,2])
```

```{r}
library(ggplot2)
ggplot(pca$x) +
  aes(PC1, PC2, color = my_cols) +
  geom_point()
```

## Digging deeper (variable loadings)

How do the original variables (i.e. the 17 different foods) contribute to our new PCs?


